Testing and Isolation Efficacy; Insights from a Simple Epidemic Model

### 2021-02-03 SIR_manuscript.tex
further updates on discussion

{\bf Delay in test reporting rate, $\omega$;}
Proposition \ref{prop1} part \ref{p1:omega} is the most interesting result. It states that returning test results more rapidly, i.e., increasing $\omega$, does not necessarily lower the reproduction number $\Rnum$; whether increasing $\omega$ lowers $\Rnum$ depends on the precise combination of model parameters (test reporting rate, testing strategies or test-specific testing weights, test sensitivity and specificity and isolation. 
Specifically, in the case of perfect isolation, i.e., when $\eta_w=0$ and $\eta_c=0$, $\Rnum$ may increase as the test reporting process become faster. This can be seen from expression \eqref{Rom}.
The interpretation for this phenomenon is that individuals are highly likely to fully self-isolate when they are awaiting test results and when they are reported positive with a negligable influence on the force of infection, $\lambda$. Thus the faster the test reporting rate, the shorter they stay in these ``\emph {safe}'' awaiting compartments, and the more they get involved in the infection process. 
Another example of this situation, when the slow test reporting rate would be favorable is in the case when the test being employed produces many false negatives. Because many infected individuals will believe they are uninfectious and thus may unknowingly spread the virus to many others.

Our model enables us to quantify the amount of delay required in the test reporting process as a strategy to reduce $\Rnum$.
To give a biological interpretation, we describe the qualitative trends predicted by the inequality \eqref{eq:necsuf} \todo{Fady please confirm}. To summarise, returning test results more rapidly tends to be more favorable, i.e., reduces $\Rnum$, when (i) \emph{Confirmed-positive} individuals, lower their contact much more than individuals who are waiting for their test results (i.e., $\eta_c \ll \eta_w$), (ii) the test is highly sensitive, i.e., $p_I$ is close to 1, and (iii) the targeted testing strategy is practiced.

Our model also predicts an asymmetry between cheap and expensive tests: if a cheap test can identify on average more infected individuals as an expensive test, then our model predicts that the cheap test will lower $\Rnum$ more. In contrast, if an expensive test can identify on average more infected individuals, it will not necessarily lower $\Rnum$ more than the cheap test. The use of tests cheaper than RT-PCR has been proposed as a potential strategy for containing the COVID-19 pandemic. While cheaper tests may be less sensitive and reliable than RT-PCR, they allow for broader and more intense testing. Using our Taylor approximation of $\Rnum$ near $\rho = 0$, we examined what circumstances (i.e., model parameters) make the use of one test more favourable than another, and give a complete description through inequality \ref{eq:rho1vsrho2}. In general, we found that the expensive test tends to more effectively lower $\Rnum$ when (a) individuals who test positive self-isolate much more than individuals who are waiting for their test result, (b) the time it takes to return tests is much shorter than the mean infectious period, and (c) the testing intensity is much greater for infected individuals than susceptible individuals.

{\bf Other points [?];}
one point we can make when discussing the relative weight of testing in different compartments is to discuss pre-testing screening tools, such as surveys or questionnaires. If we have a quantitative description of how the testing intensities affect the dynamics, we can make statements like ``our results suggest that employing a pre-testing screening tool can help target infected individuals more effectively. In particular, doubling the sensitivity of the pre-screening tool would *do something* to $\Rnum$.

% JD:
% \begin{itemize}
%  \item We're basically looking at the potential advantage of slow tests:
% people waiting for test results may be more careful than those who
% have received negatives. These advantages are real, and neglected. We
% also compare to an individual-level advantage of fast tests: people
% who test positive may be even more careful. But we are missing out on
% community-level advantages of fast testing: better assessment of the
% situation, identification of hot spots, contact-tracing, etc. 
% \item  Maybe something about the testing rate/intensity, now the per capita testing intensity is very low ($\rho \approx 0$). In near future new test kits may be widely accessible, our model provide insights in this case (what are those?)
% \item ``delay negative results" result
% \end{itemize}

>>>>

{\bf The potential advantage of slow test reporting, or favorable-delay-reporting;}
Individuals are highly likely to fully self-isolate when they are either awaiting the test results or they are reported, thus reducing or eliminating their potentially infectious contacts. Thus, the faster the test reporting rate, $\omega$, the shorter these individuals stay in the ``\emph {safe}'' awaiting-confirmed compartments, namely $Z_n$, $Z_p$ and $Z_c$ for $Z \in \{S,I,R\}$, and the more they get involved in the infection process.
This advantage of slow test reporting is real, and neglected. 
We also compare to an individual-level advantage of fast tests: people who test positive may be even more careful.
In our model analysis, Proposition \ref{prop1} part \ref{p1:omega} is describing this potential advantage. 
It states that returning test results more rapidly, i.e., increasing the rate $\omega$, does not necessarily lower the reproduction number $\Rnum$; whether increasing $\omega$ lowers $\Rnum$ depends on the precise combination of model parameters  including test reporting rate, testing strategies represented by compartment-specific testing weights, test sensitivity and specificity, and the level of isolation. 
Specifically, in the case of perfect isolation, i.e., when $\eta_w=0$ and $\eta_c=0$, $\Rnum$ may increase as the test reporting process becomes faster. This can be seen from expression \eqref{Rom}.
Another example of this favorable delay in $\omega$ could be when the test being employed produces many false negatives. Because many infected asymptomatic individuals will believe they are uninfected/uninfectious, thus may unknowingly spread the virus to many others. Agian the delay in the test reporting rate keeps these individuals in the ``\emph {safe}'' awaiting-confirmed compartments.

we are missing out on
community-level advantages of fast testing: better assessment of the
situation, identification of hot spots, contact-tracing, etc.
The \percap testing intensity of CIVID-19 after about a year from the first case reported in December 2019, is still low (($\rho \approx 0$) in our model). In near future new test kits may be widely accessible, our model provide insights in this case. 
In particular, if a cheap test can identify on average more infected individuals as an expensive test, then our model predicts that the cheap test will lower $\Rnum$ more. In contrast, if an expensive test can identify on average more infected individuals, it will not necessarily lower $\Rnum$ more than the cheap test. The use of tests cheaper than RT-PCR has been proposed as a potential strategy for containing the COVID-19 pandemic. While cheaper tests may be less sensitive and reliable than RT-PCR, they allow for broader and more intense testing. Using our Taylor approximation of $\Rnum$ near $\rho = 0$, we examined what circumstances (i.e., model parameters) make the use of one test more favourable than another, and give a complete description through inequality \ref{eq:rho1vsrho2}. In general, we found that the expensive test tends to more effectively lower $\Rnum$ when (a) individuals who test positive self-isolate much more than individuals who are waiting for their test result, (b) the time it takes to return tests is much shorter than the mean infectious period, and (c) the testing intensity is much greater for infected individuals than susceptible individuals.

\ali{ not sure if we want these 2 following parageraphs, or shrink it into afew short sentences?}
In addition to the favorable-delay-reporting observation of our model \eqref{model}, the model enables us to quantify the amount of delay required in the test reporting process as a strategy to reduce $\Rnum$.
To give a biological interpretation, we describe the qualitative trends predicted by the inequality \eqref{eq:necsuf} \todo{Fady please confirm}. To summarise, returning test results more rapidly tends to be more favorable, i.e., reduces $\Rnum$, when (i) \emph{Confirmed-positive} individuals, lower their contact much more than individuals who are waiting for their test results (i.e., $\eta_c \ll \eta_w$), (ii) the test is highly sensitive, i.e., $p_I$ is close to 1, and (iii) the targeted testing strategy is practiced.

one point we can make when discussing the relative weight of testing in different compartments is to discuss pre-testing screening tools, such as surveys or questionnaires. If we have a quantitative description of how the testing intensities affect the dynamics, we can make statements like ``our results suggest that employing a pre-testing screening tool can help target infected individuals more effectively. In particular, doubling the sensitivity of the pre-screening tool would *do something* to $\Rnum$.
